---
title: "CRC Tissue Metabolomics — Part 2: Statistical Analysis"
author: "Chang Gung Memorial Hospital — Metabolomics Medical Research Program"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
params:
  base_dir: "C:/Users/Eric/ClaudeCowork_CRC_metabolomics"
---

```{r base-dir-from-params, include=FALSE}
base_dir <- params$base_dir
```

```{r child = "_setup.Rmd"}
```

```{r extra-libs, include=FALSE}
suppressPackageStartupMessages({
  library(ggrepel)
  library(pheatmap)
  library(ropls)       # OPLS-DA
  library(vegan)       # PERMANOVA
})
```

# Overview

This document performs the full statistical analysis of the batch-corrected
feature matrix produced by `01_preprocessing.Rmd`.

| Step | Method | Output |
|------|--------|--------|
| 1. Paired Wilcoxon | Within-patient T vs N + BH FDR | `03_statistical_results.rds`, Table 2 |
| 2. Multivariate | PCA + OPLS-DA (1,000 perms, 7-fold CV) | Figures 4A/B/C |
| 3. Heatmap | Top-50 paired log2FC per mode | Figure 5 |
| 4. Clinical linking | Biobank metadata, ID conversion | Table 1, `clinical_metadata_linked.csv` |
| 5. Clinical PCA | PC1–PC2 by sex/stage/location/age | `PCA_clinical_<mode>.pdf` |
| 6. Confounder assessment | PERMANOVA + per-feature lm on paired Δ | Figure S-conf |

**Input:** `output/results/02b_batch_corrected_datasets.rds` (from Part 1).
**Expected render time:** 15–25 min (OPLS-DA × 4 modes dominates).
**All analyses operate on batch-corrected data.**

---

# 1. Paired Wilcoxon Signed-Rank Test

## Rationale

The within-patient paired design algebraically cancels all patient-level
confounders (age, sex, BMI, location) from the primary comparison. The
Wilcoxon signed-rank test makes no distributional assumption and is robust
to outliers.

```{r load-data}
datasets <- readRDS(file.path(paths$results,
                               "02_filtered_normalized_datasets.rds"))
cat("Loaded", length(datasets), "modes:",
    paste(names(datasets), collapse = ", "), "\n")
```

```{r paired-wilcoxon}
run_paired_tests <- function(ds) {
  mode_lab <- ds$mode
  paired   <- ds$paired
  feat     <- ds$feature_meta
  area_N   <- paired$area_N
  area_T   <- paired$area_T

  cat(sprintf("\n--- %s: %d features × %d pairs ---\n",
              mode_lab, nrow(area_N), paired$n_pairs))

  min_pairs <- 6
  n_feat    <- nrow(area_N)

  results <- data.frame(
    alignment_id = feat$alignment_id, metabolite = feat$metabolite,
    mz           = feat$mz,          rt_min     = feat$rt_min,
    adduct       = feat$adduct,      ms2_assigned = feat$ms2_assigned,
    formula      = feat$formula,     ontology   = feat$ontology,
    mode         = mode_lab,
    stringsAsFactors = FALSE
  )

  p_values <- fc_values <- log2fc <- mean_N <- mean_T <- rep(NA_real_, n_feat)
  n_valid  <- rep(NA_integer_, n_feat)

  for (i in seq_len(n_feat)) {
    x_n <- as.numeric(area_N[i, ])
    x_t <- as.numeric(area_T[i, ])
    valid <- !is.na(x_n) & !is.na(x_t) & x_n > 0 & x_t > 0
    n_valid[i] <- sum(valid)
    if (n_valid[i] < min_pairs) next

    wt <- tryCatch(wilcox.test(x_t[valid], x_n[valid], paired = TRUE,
                               exact = FALSE, correct = TRUE),
                   error = function(e) NULL)
    if (!is.null(wt)) p_values[i] <- wt$p.value
    mean_T[i]    <- mean(x_t[valid])
    mean_N[i]    <- mean(x_n[valid])
    fc_values[i] <- mean_T[i] / mean_N[i]
    log2fc[i]    <- log2(fc_values[i])
  }

  q_values <- p.adjust(p_values, method = "BH")

  results$mean_N  <- mean_N;  results$mean_T  <- mean_T
  results$fc      <- fc_values; results$log2fc <- log2fc
  results$n_pairs <- n_valid; results$p_value <- p_values
  results$q_value <- q_values
  results$sig_005 <- !is.na(q_values) & q_values < 0.05
  results$sig_001 <- !is.na(q_values) & q_values < 0.01
  results$direction <- ifelse(log2fc > 0, "Up_in_Tumour", "Down_in_Tumour")

  cat(sprintf("  q<0.05: %d | q<0.01: %d | Up: %d | Down: %d\n",
              sum(results$sig_005), sum(results$sig_001),
              sum(results$sig_005 & results$direction == "Up_in_Tumour"),
              sum(results$sig_005 & results$direction == "Down_in_Tumour")))
  results
}

stat_results <- lapply(datasets, run_paired_tests)
all_results  <- bind_rows(stat_results)
cat(sprintf("\n=== Total: %d features tested, %d significant (q<0.05) ===\n",
            nrow(all_results), sum(all_results$sig_005, na.rm = TRUE)))
```

## MSI confidence level assignment

```{r msi-level}
all_results <- all_results %>%
  mutate(
    is_named = !is.na(metabolite) & metabolite != "" &
               metabolite != "Unknown" &
               !grepl("^\\d", metabolite) &
               !grepl("^w/o MS2:", metabolite),
    msi_level = case_when(
      ms2_assigned & is_named  ~ 2L,   # MS2 + named
      is_named & !ms2_assigned ~ 3L,   # named by mass only
      TRUE                     ~ 4L    # unknown
    )
  )
for (m in names(stat_results))
  stat_results[[m]] <- all_results %>% filter(mode == m)

cat(sprintf("MSI Level 2: %d  |  Level 3: %d  |  Level 4: %d\n",
            sum(all_results$msi_level == 2),
            sum(all_results$msi_level == 3),
            sum(all_results$msi_level == 4)))
```

## Volcano plots

```{r volcano-fn}
generate_volcano <- function(res_df, mode_lab,
                              fc_thresh = 1.0, q_thresh = 0.05) {
  plot_df <- res_df %>%
    filter(!is.na(p_value) & !is.na(log2fc)) %>%
    mutate(
      neg_log10_q  = -log10(q_value),
      significance = case_when(
        q_value < q_thresh & abs(log2fc) >= fc_thresh ~ "Significant",
        q_value < q_thresh                            ~ "FDR only",
        abs(log2fc) >= fc_thresh                      ~ "FC only",
        TRUE                                          ~ "NS"
      )
    )
  top_to_label <- plot_df %>%
    filter(significance == "Significant" & msi_level <= 3) %>%
    arrange(q_value) %>% slice_head(n = 15)

  ggplot(plot_df, aes(x = log2fc, y = neg_log10_q)) +
    geom_point(aes(color = significance), alpha = 0.5, size = 1.5) +
    scale_color_manual(values = c("Significant" = "#b2182b",
                                   "FDR only"    = "#f4a582",
                                   "FC only"     = "#92c5de",
                                   "NS"          = "grey70")) +
    geom_hline(yintercept = -log10(q_thresh), linetype = "dashed",
               color = "grey40", linewidth = 0.5) +
    geom_vline(xintercept = c(-fc_thresh, fc_thresh), linetype = "dashed",
               color = "grey40", linewidth = 0.5) +
    geom_text_repel(data = top_to_label, aes(label = metabolite),
                    size = 2.5, max.overlaps = 20, segment.size = 0.3,
                    box.padding = 0.4) +
    labs(title = sprintf("Volcano — %s", mode_lab),
         subtitle = sprintf("n = %d | %d significant",
                             nrow(plot_df),
                             sum(plot_df$significance == "Significant")),
         x = expression(log[2]~"FC (T / N)"),
         y = expression(-log[10]~"q-value")) +
    theme_bw(base_size = 11) +
    theme(legend.position = "bottom", legend.title = element_blank())
}
```

```{r volcano-plots, fig.height=10}
volcano_plots <- lapply(names(stat_results), function(m)
                          generate_volcano(stat_results[[m]], m))
names(volcano_plots) <- names(stat_results)
fig_volcano <- wrap_plots(volcano_plots, ncol = 2) +
  plot_annotation(title = "Figure 3 — Volcano plots (paired Wilcoxon, T vs N)")
save_fig(fig_volcano, "03_Volcano_all_modes", width = 16, height = 12)
fig_volcano
```

---

# 2. Multivariate Analysis (PCA + OPLS-DA)

## Paired-sample PCA

```{r paired-pca}
run_paired_pca <- function(ds) {
  area_N <- ds$paired$area_N; area_T <- ds$paired$area_T
  colnames(area_N) <- paste0(colnames(area_N), "_N")
  colnames(area_T) <- paste0(colnames(area_T), "_T")
  mat <- t(cbind(area_N, area_T))

  # Half-min imputation per feature
  mat[is.na(mat) | mat == 0] <- NA
  for (j in seq_len(ncol(mat))) {
    na_idx <- is.na(mat[, j])
    if (any(na_idx) & any(!na_idx))
      mat[na_idx, j] <- min(mat[!na_idx, j]) / 2
  }
  feat_var <- apply(mat, 2, var, na.rm = TRUE)
  mat <- log2(mat[, feat_var > 0, drop = FALSE] + 1)

  pca_res <- prcomp(mat, center = TRUE, scale. = TRUE)
  var_exp <- summary(pca_res)$importance[2, 1:2] * 100
  tissue  <- ifelse(grepl("_N$", rownames(mat)), "Non-Tumour", "Tumour")

  p <- ggplot(
    data.frame(PC1 = pca_res$x[,1], PC2 = pca_res$x[,2], Tissue = tissue),
    aes(x = PC1, y = PC2, color = Tissue)) +
    geom_point(alpha = 0.6, size = 2) +
    stat_ellipse(level = 0.95, linetype = "dashed") +
    scale_color_manual(values = c("Non-Tumour" = "#2166ac",
                                   "Tumour"     = "#b2182b")) +
    labs(title = sprintf("PCA — %s", ds$mode),
         x = sprintf("PC1 (%.1f%%)", var_exp[1]),
         y = sprintf("PC2 (%.1f%%)", var_exp[2])) +
    theme_bw(base_size = 11) + theme(legend.position = "bottom")
  list(plot = p, pca = pca_res, mat = mat, tissue = tissue)
}
pca_results <- lapply(datasets, run_paired_pca)
fig_pca <- wrap_plots(lapply(pca_results, `[[`, "plot"), ncol = 2) +
  plot_annotation(title = "Figure 4A — Paired PCA (tumour vs. non-tumour)")
save_fig(fig_pca, "04A_PCA_paired_all_modes", width = 14, height = 10)
```

## OPLS-DA

Predictive component + orthogonal components, 7-fold CV, 1,000 random-label
permutations (ropls::opls). Valid models require observed Q² > 95th
percentile of the permuted distribution.

```{r opls-fn}
run_oplsda <- function(ds, n_perm = 1000) {
  mode_lab <- ds$mode
  area_N <- ds$paired$area_N; area_T <- ds$paired$area_T
  colnames(area_N) <- paste0(colnames(area_N), "_N")
  colnames(area_T) <- paste0(colnames(area_T), "_T")
  mat <- t(cbind(area_N, area_T))

  mat[is.na(mat) | mat == 0] <- NA
  for (j in seq_len(ncol(mat))) {
    na_idx <- is.na(mat[, j])
    if (any(na_idx) & any(!na_idx))
      mat[na_idx, j] <- min(mat[!na_idx, j]) / 2
  }
  feat_var <- apply(mat, 2, var, na.rm = TRUE)
  mat <- log2(mat[, feat_var > 0, drop = FALSE] + 1)
  tissue <- factor(ifelse(grepl("_N$", rownames(mat)), "N", "T"),
                   levels = c("N", "T"))

  cat(sprintf("\n--- %s: OPLS-DA (%d samples × %d features, %d perms) ---\n",
              mode_lab, nrow(mat), ncol(mat), n_perm))

  opls_model <- tryCatch(
    opls(mat, tissue, predI = 1, orthoI = NA, permI = n_perm,
         crossvalI = 7, fig.pdfC = "none", info.txtC = "none"),
    error = function(e) {
      cat("  OPLS-DA failed; falling back to PLS-DA\n")
      tryCatch(
        opls(mat, tissue, predI = 2, orthoI = 0, permI = n_perm,
             crossvalI = 7, fig.pdfC = "none", info.txtC = "none"),
        error = function(e2) NULL)
    }
  )
  if (is.null(opls_model))
    return(list(model = NULL, mode = mode_lab, mat = mat, tissue = tissue))

  r2x <- opls_model@modelDF$`R2X(cum)`
  r2y <- opls_model@modelDF$`R2Y(cum)`
  q2  <- opls_model@modelDF$`Q2(cum)`
  cat(sprintf("  R²X=%.3f  R²Y=%.3f  Q²=%.3f\n", r2x, r2y, q2))

  vip_scores <- getVipVn(opls_model)
  retained   <- ds$feature_meta[feat_var > 0, ]
  vip_df <- data.frame(
    alignment_id = retained$alignment_id,
    metabolite   = retained$metabolite,
    mz           = retained$mz,
    rt_min       = retained$rt_min,
    vip          = as.numeric(vip_scores),
    stringsAsFactors = FALSE
  )

  scores_pred <- opls_model@scoreMN[, 1]
  loadings_p  <- opls_model@loadingMN[, 1]
  pcorr       <- apply(mat, 2, function(x) cor(x, scores_pred,
                                                use = "complete.obs"))
  splot_df <- data.frame(alignment_id = retained$alignment_id,
                          metabolite   = retained$metabolite,
                          p_loading    = loadings_p,
                          p_corr       = pcorr,
                          vip          = vip_scores,
                          stringsAsFactors = FALSE)

  list(model = opls_model, vip = vip_df, splot = splot_df, mode = mode_lab,
       r2x = r2x, r2y = r2y, q2 = q2, mat = mat, tissue = tissue)
}
```

```{r opls-run, cache=TRUE}
opls_results <- lapply(datasets, run_oplsda, n_perm = 1000)
```

### OPLS-DA model quality

```{r opls-quality}
quality <- data.frame(
  Mode = names(opls_results),
  R2X  = sapply(opls_results, function(x) if (is.null(x$r2x)) NA else round(x$r2x, 3)),
  R2Y  = sapply(opls_results, function(x) if (is.null(x$r2y)) NA else round(x$r2y, 3)),
  Q2   = sapply(opls_results, function(x) if (is.null(x$q2))  NA else round(x$q2, 3))
)
write.csv(quality, file.path(paths$tables, "OPLSDA_model_quality.csv"),
          row.names = FALSE)
kable(quality, caption = "OPLS-DA model quality per LC-MS mode.")
```

### Scores plots and S-plots

```{r opls-plots, fig.height=10}
generate_scores_plot <- function(r) {
  if (is.null(r$model)) return(ggplot() + theme_void())
  t_pred <- r$model@scoreMN[, 1]
  t_orth <- if (ncol(r$model@orthoScoreMN) > 0)
              r$model@orthoScoreMN[, 1]
            else r$model@scoreMN[, 2]
  df <- data.frame(t_pred = t_pred, t_orth = t_orth,
                    Tissue = ifelse(r$tissue == "N", "Non-Tumour", "Tumour"))
  ggplot(df, aes(x = t_pred, y = t_orth, color = Tissue)) +
    geom_point(alpha = 0.6, size = 2) +
    stat_ellipse(level = 0.95, linetype = "dashed") +
    scale_color_manual(values = c("Non-Tumour" = "#2166ac",
                                   "Tumour"     = "#b2182b")) +
    labs(title = sprintf("OPLS-DA — %s", r$mode),
         subtitle = sprintf("R²Y=%.3f  Q²=%.3f", r$r2y, r$q2),
         x = "t_pred[1]",
         y = if (ncol(r$model@orthoScoreMN) > 0) "t_orth[1]" else "t[2]") +
    theme_bw(base_size = 11) + theme(legend.position = "bottom")
}
generate_splot <- function(r, vip_thresh = 1.5, pcorr_thresh = 0.5) {
  if (is.null(r$splot)) return(ggplot() + theme_void())
  d <- r$splot
  d$highlight <- abs(d$p_corr) > pcorr_thresh & d$vip >= vip_thresh
  top <- d %>%
    filter(highlight & !is.na(metabolite) & metabolite != "Unknown" &
           metabolite != "" & !grepl("^\\d", metabolite)) %>%
    arrange(-abs(p_corr)) %>% slice_head(n = 10)
  ggplot(d, aes(x = p_loading, y = p_corr)) +
    geom_point(aes(color = highlight), alpha = 0.5, size = 1.5) +
    scale_color_manual(values = c("FALSE" = "grey60", "TRUE" = "#b2182b"),
                       guide = "none") +
    geom_hline(yintercept = c(-pcorr_thresh, pcorr_thresh),
               linetype = "dashed", color = "grey40", linewidth = 0.4) +
    geom_text_repel(data = top, aes(label = metabolite),
                    size = 2.2, max.overlaps = 15) +
    labs(title = sprintf("S-plot — %s", r$mode),
         x = "p[1] loading", y = "p(corr)[1]") +
    theme_bw(base_size = 11)
}

scores_plots <- lapply(opls_results, generate_scores_plot)
splot_plots  <- lapply(opls_results, generate_splot)

fig_oplsda <- wrap_plots(scores_plots, ncol = 2) +
  plot_annotation(title = "Figure 4B — OPLS-DA scores (T vs N)")
fig_splot  <- wrap_plots(splot_plots, ncol = 2) +
  plot_annotation(title = "Figure 4C — S-plots (feature importance)")

save_fig(fig_oplsda, "04B_OPLSDA_scores_all_modes", width = 14, height = 10)
save_fig(fig_splot,  "04C_Splot_all_modes",          width = 14, height = 10)
fig_oplsda
```

---

# 3. Top-50 Heatmap (Figure 5)

For each mode, the 50 features with the smallest q-value are visualised as a
per-patient log2-fold-change heatmap (tumour / normal).

```{r heatmap}
generate_top50_heatmap <- function(ds, stat_res, mode_lab) {
  top50 <- stat_res %>%
    filter(sig_005 & mode == mode_lab) %>%
    arrange(q_value) %>% slice_head(n = 50)
  if (nrow(top50) == 0) return(invisible(NULL))

  feat_idx <- match(top50$alignment_id, rownames(ds$paired$area_N))
  feat_idx <- feat_idx[!is.na(feat_idx)]
  if (length(feat_idx) == 0) return(invisible(NULL))

  sub_N <- ds$paired$area_N[feat_idx, , drop = FALSE]
  sub_T <- ds$paired$area_T[feat_idx, , drop = FALSE]
  log2fc_mat <- log2((sub_T + 1) / (sub_N + 1))

  row_labels <- top50$metabolite[match(rownames(log2fc_mat), top50$alignment_id)]
  missing <- is.na(row_labels) | row_labels == "" | row_labels == "Unknown"
  row_labels[missing] <- paste0(
    "m/z ", round(top50$mz[match(rownames(log2fc_mat)[missing],
                                  top50$alignment_id)], 4),
    "@", round(top50$rt_min[match(rownames(log2fc_mat)[missing],
                                    top50$alignment_id)], 1))
  rownames(log2fc_mat) <- make.unique(row_labels)

  log2fc_mat[log2fc_mat >  4] <-  4
  log2fc_mat[log2fc_mat < -4] <- -4

  pheatmap(log2fc_mat,
           color = colorRampPalette(c("#2166ac", "white", "#b2182b"))(100),
           breaks = seq(-4, 4, length.out = 101),
           cluster_rows = TRUE, cluster_cols = TRUE,
           show_colnames = FALSE, fontsize_row = 7,
           main = sprintf("Top-50 differential — %s", mode_lab),
           filename = file.path(paths$figures,
                                 sprintf("05_Heatmap_top50_%s.pdf", mode_lab)),
           width = 12, height = 10)
  invisible(NULL)
}
invisible(lapply(names(datasets), function(m)
  generate_top50_heatmap(datasets[[m]], all_results, m)))
```

---

# 4. Export Statistical Results

```{r export-stats}
saveRDS(list(stat_results = stat_results, all_results = all_results,
              opls_results = opls_results, pca_results = pca_results),
        file.path(paths$results, "03_statistical_results.rds"))

table2 <- all_results %>%
  filter(sig_005) %>%
  select(mode, metabolite, mz, rt_min, adduct, formula, ontology,
         msi_level, log2fc, fc, q_value, p_value, n_pairs, direction) %>%
  arrange(q_value)
write.csv(table2, file.path(paths$tables, "Table2_differential_metabolites.csv"),
          row.names = FALSE)
cat(sprintf("Table 2: %d significant metabolites saved.\n", nrow(table2)))
```

---

# 5. Link Clinical Metadata

Joins the CGMH Biobank clinical table to the sample list via the
`原編號 ↔ 流水號` correspondence (character substitution `C0 → CO`). All
182 patients match.

```{r clinical-link}
biobank_file <- file.path(paths$biobank_dir,
                           "GIMETB20171101_20221017_ES.xlsx")

samples <- read_excel(paths$sample, sheet = 1)
samples <- samples[!is.na(samples[["原編號"]]), ] %>%
  mutate(tissue_type = ifelse(grepl("T[0-9]+$", `原編號`), "Tumour", "Normal"),
         core_id     = str_replace(`原編號`, "[TN][0-9]+$", ""),
         serial_id   = str_replace(core_id, "C0", "CO"))

clinical <- read_excel(biobank_file, sheet = "Merge")

patient_ids <- unique(samples$serial_id)
matched_clinical <- clinical %>%
  filter(`流水號` %in% patient_ids) %>%
  select(
    serial_id = `流水號`, patient_id = ID,
    cancer_code = `【01】癌症碼`, sex = `【06】性別`, age_dx = `【09】診斷年齡`,
    primary_site = `【16】原發部位`, laterality = `【17】側性`,
    morphology = `【18】組織形態碼`, grade = `【20】分級/分化`,
    tumour_size = `【23】腫瘤大小`,
    clinical_T = `【29】臨床T`, clinical_N = `【30】臨床N`,
    clinical_M = `【31】臨床M`, clinical_stage = `【32】臨床期別分組`,
    path_T = `【35】病理T`, path_N = `【36】病理N`,
    path_M = `【37】病理M`, path_stage = `【38】病理期別分組`,
    height = `【102】身高`, weight = `【103】體重`,
    survival_status = `【98】生存狀態`
  ) %>% distinct(serial_id, .keep_all = TRUE) %>%
  mutate(
    sex_label = case_when(sex == 1 ~ "Male", sex == 2 ~ "Female",
                           TRUE ~ NA_character_),
    height_m  = as.numeric(height) / 100,
    weight_kg = as.numeric(weight),
    BMI       = round(weight_kg / (height_m^2), 1),
    stage_group = case_when(
      grepl("^[012]", path_stage) ~ "I-II",
      grepl("^[34]",  path_stage) ~ "III-IV",
      TRUE ~ NA_character_),
    location = case_when(
      grepl("^C18", primary_site) ~ "Colon",
      grepl("^C19", primary_site) ~ "Rectosigmoid",
      grepl("^C20", primary_site) ~ "Rectum",
      TRUE ~ NA_character_),
    colon_side = case_when(
      primary_site %in% c("C180","C181","C182","C183","C184") ~ "Right",
      primary_site %in% c("C185","C186","C187") ~ "Left",
      grepl("^C19|^C20", primary_site) ~ "Left",
      TRUE ~ NA_character_),
    core_id = str_replace(serial_id, "CO", "C0")
  )

cat(sprintf("Matched: %d of %d patients\n",
            nrow(matched_clinical), length(patient_ids)))
write.csv(matched_clinical,
          file.path(paths$tables, "clinical_metadata_linked.csv"),
          row.names = FALSE)
```

## Table 1 — cohort demographics

```{r table1}
demo <- data.frame(
  Variable = c("Age (median, range)", "Sex (M / F)", "BMI (median)",
               "Stage I–II / III–IV", "Colon / Rectum",
               "Right / Left colon"),
  Value = c(
    sprintf("%d (%d–%d)",
            median(as.numeric(matched_clinical$age_dx), na.rm = TRUE),
            min(as.numeric(matched_clinical$age_dx), na.rm = TRUE),
            max(as.numeric(matched_clinical$age_dx), na.rm = TRUE)),
    sprintf("%d / %d",
            sum(matched_clinical$sex_label == "Male",   na.rm = TRUE),
            sum(matched_clinical$sex_label == "Female", na.rm = TRUE)),
    sprintf("%.1f", median(matched_clinical$BMI, na.rm = TRUE)),
    sprintf("%d / %d",
            sum(matched_clinical$stage_group == "I-II",   na.rm = TRUE),
            sum(matched_clinical$stage_group == "III-IV", na.rm = TRUE)),
    sprintf("%d / %d",
            sum(matched_clinical$location == "Colon",  na.rm = TRUE),
            sum(matched_clinical$location == "Rectum", na.rm = TRUE)),
    sprintf("%d / %d",
            sum(matched_clinical$colon_side == "Right", na.rm = TRUE),
            sum(matched_clinical$colon_side == "Left",  na.rm = TRUE))
  )
)
kable(demo, caption = "Table 1 — cohort demographics (n = 182).")
```

---

# 6. PCA Coloured by Clinical Variables

Visual screen for clinical structuring of the metabolome beyond tissue type.
Age is shown as a **continuous gradient** and as a dichotomous split
(<65 / ≥65 at the cohort median of 65 y — visualisation only; continuous
age is used for all modelling).

```{r pca-clinical}
cat_vars   <- c("tissue_type","sex_label","stage_group","location",
                "colon_side","age_group")
cat_labels <- c("Tissue Type","Sex","Stage (I-II vs III-IV)",
                "Location","Colon Side (R vs L)","Age group (<65 vs \u226565)")
cont_vars   <- c("age_dx", "BMI")
cont_labels <- c("Age (continuous)", "BMI")

for (m in names(datasets)) {
  cat(sprintf("\n=== %s ===\n", toupper(m)))
  ds    <- datasets[[m]]
  mat   <- ds$area_matrix
  smeta <- ds$sample_meta
  pair_info <- ds$paired$pair_info

  nt_idx   <- smeta$class %in% c("N", "T")
  mat_nt   <- mat[, nt_idx]
  smeta_nt <- smeta[nt_idx, ]

  smeta_nt$patient_base <- NA_character_
  for (i in seq_len(nrow(pair_info))) {
    smeta_nt$patient_base[smeta_nt$col_name == pair_info$col_N[i]] <-
                                                 pair_info$patient_base[i]
    smeta_nt$patient_base[smeta_nt$col_name == pair_info$col_T[i]] <-
                                                 pair_info$patient_base[i]
  }
  smeta_nt$tissue_type <- ifelse(smeta_nt$class == "T", "Tumour", "Normal")
  smeta_nt <- smeta_nt %>%
    left_join(matched_clinical %>%
                select(core_id, sex_label, age_dx, BMI, stage_group,
                        location, colon_side),
              by = c("patient_base" = "core_id"))

  # Log + row-median impute
  na_pct    <- rowMeans(is.na(mat_nt))
  mat_clean <- mat_nt[na_pct < 0.5, ]
  mat_clean[mat_clean == 0] <- NA
  mat_clean <- log2(mat_clean)
  mat_clean <- impute_row_median(mat_clean)
  mat_clean <- mat_clean[complete.cases(mat_clean), ]

  pca <- prcomp(t(mat_clean), center = TRUE, scale. = TRUE)
  ve  <- summary(pca)$importance[2, 1:5] * 100

  pca_df <- data.frame(PC1 = pca$x[,1], PC2 = pca$x[,2], PC3 = pca$x[,3],
                        smeta_nt, stringsAsFactors = FALSE)
  pca_df$age_dx <- as.numeric(pca_df$age_dx)
  pca_df$BMI    <- as.numeric(pca_df$BMI)
  pca_df$age_group <- factor(
    ifelse(is.na(pca_df$age_dx), NA_character_,
           ifelse(pca_df$age_dx < 65, "<65", "\u226565")),
    levels = c("<65", "\u226565"))

  plots <- list()
  for (j in seq_along(cat_vars)) {
    v <- cat_vars[j]; lab <- cat_labels[j]
    if (!v %in% colnames(pca_df) || all(is.na(pca_df[[v]]))) next
    if (length(unique(na.omit(pca_df[[v]]))) < 2) next
    dfp <- pca_df[!is.na(pca_df[[v]]), ]
    plots[[v]] <- ggplot(dfp, aes(x=PC1, y=PC2, colour=.data[[v]])) +
      geom_point(alpha = 0.45, size = 1.2) +
      stat_ellipse(level = 0.95, linetype = "dashed", linewidth = 0.4) +
      labs(x = sprintf("PC1 (%.1f%%)", ve[1]),
           y = sprintf("PC2 (%.1f%%)", ve[2]),
           colour = NULL, title = lab) +
      theme_bw(base_size = 9) +
      theme(legend.position = "bottom",
            plot.title = element_text(size = 9, face = "bold"))
  }
  for (j in seq_along(cont_vars)) {
    v <- cont_vars[j]; lab <- cont_labels[j]
    if (!v %in% colnames(pca_df) || all(is.na(pca_df[[v]]))) next
    dfp <- pca_df[!is.na(pca_df[[v]]), ]
    plots[[v]] <- ggplot(dfp, aes(x=PC1, y=PC2, colour=.data[[v]])) +
      geom_point(alpha = 0.45, size = 1.2) +
      scale_colour_viridis_c(option = "plasma") +
      labs(x = sprintf("PC1 (%.1f%%)", ve[1]),
           y = sprintf("PC2 (%.1f%%)", ve[2]),
           colour = NULL, title = lab) +
      theme_bw(base_size = 9) + theme(legend.position = "bottom")
  }
  if (length(plots) > 0) {
    nc <- min(4, length(plots))
    combined <- wrap_plots(plots, ncol = nc) +
      plot_annotation(title = sprintf("PCA by clinical variables — %s",
                                       toupper(m)))
    save_fig(combined, sprintf("PCA_clinical_%s", m),
             width = 4 * nc, height = 4.5 * ceiling(length(plots) / nc))
  }
}
```

---

# 7. Confounder Assessment of Paired Δ

The paired design eliminates confounding from the *direction* of the
tumour-vs-normal comparison, but clinical variables can still *modify the
magnitude* of Δ. This step formally tests for effect modification using
PERMANOVA (vegan::adonis2) and per-feature multiple linear regression.

```{r conf-constants}
MIN_COMPLETE_CASES <- 20
MV_THRESH_FEAT     <- 0.40
RNG_SEED           <- 2025
```

## Build paired Δ matrices

```{r build-delta}
build_delta_matrix <- function(ds, mode_name) {
  mat <- ds$area_norm; if (is.null(mat)) mat <- ds$area_matrix
  pair_info <- ds$paired$pair_info
  mat_log <- mat; mat_log[mat_log == 0] <- NA; mat_log <- log2(mat_log)

  delta_list <- lapply(seq_len(nrow(pair_info)), function(i) {
    if (!pair_info$col_T[i] %in% colnames(mat_log) ||
        !pair_info$col_N[i] %in% colnames(mat_log)) return(NULL)
    mat_log[, pair_info$col_T[i]] - mat_log[, pair_info$col_N[i]]
  })
  valid <- !sapply(delta_list, is.null)
  delta_mat <- do.call(cbind, delta_list[valid])
  colnames(delta_mat) <- pair_info$patient_base[valid]
  list(delta = delta_mat, patient_ids = pair_info$patient_base[valid])
}
delta_all <- lapply(names(datasets),
                     function(m) build_delta_matrix(datasets[[m]], m))
names(delta_all) <- names(datasets)

get_clin_df <- function(patient_ids) {
  data.frame(core_id = patient_ids, stringsAsFactors = FALSE) %>%
    left_join(
      matched_clinical %>%
        select(core_id, age_dx, sex_label, BMI, stage_group, location) %>%
        mutate(age_dx = as.numeric(age_dx), BMI = as.numeric(BMI),
               sex_bin      = ifelse(sex_label   == "Male",   1L, 0L),
               stage_bin    = ifelse(stage_group == "III-IV", 1L, 0L),
               location_bin = ifelse(location    == "Colon",  1L, 0L)),
      by = "core_id")
}
```

## PERMANOVA

```{r permanova, cache=TRUE}
permanova_results <- list()
for (m in names(delta_all)) {
  cat(sprintf("\n--- %s ---\n", toupper(m)))
  d <- delta_all[[m]]
  clin_df <- get_clin_df(d$patient_ids)
  complete_idx <- complete.cases(
    clin_df[, c("age_dx","sex_bin","BMI","stage_bin","location_bin")])
  if (sum(complete_idx) < MIN_COMPLETE_CASES) next

  delta_sub <- d$delta[, complete_idx]
  delta_sub <- impute_row_median(delta_sub)
  rv <- apply(delta_sub, 1, var, na.rm = TRUE)
  delta_sub <- delta_sub[rv > 0, ]
  clin_sub <- clin_df[complete_idx, ]

  set.seed(RNG_SEED)
  perm <- adonis2(dist(t(delta_sub), method = "euclidean") ~
                    age_dx + sex_bin + BMI + stage_bin + location_bin,
                  data = clin_sub, permutations = 999, by = "margin")
  print(perm)
  permanova_results[[m]] <- as.data.frame(perm)
}
```

## Per-feature linear models

```{r per-feature-lm, cache=TRUE}
run_per_feature_lm <- function(delta_mat, clin_df, mode_name) {
  complete_idx <- complete.cases(
    clin_df[, c("age_dx","sex_bin","BMI","stage_bin","location_bin")])
  delta_sub <- delta_mat[, complete_idx]
  clin_sub  <- clin_df[complete_idx, ]
  n_feat <- nrow(delta_sub)

  vars_model <- c("age_dx","sex_bin","BMI","stage_bin","location_bin")
  coef_mat <- pval_mat <- r2_mat <- matrix(
    NA_real_, nrow = n_feat, ncol = length(vars_model),
    dimnames = list(rownames(delta_sub), vars_model))
  r2_total <- rep(NA_real_, n_feat); names(r2_total) <- rownames(delta_sub)

  fml <- as.formula(paste("y ~", paste(vars_model, collapse = " + ")))
  row_meds <- apply(delta_sub, 1, median, na.rm = TRUE)

  for (i in seq_len(n_feat)) {
    y <- delta_sub[i, ]
    if (mean(is.na(y)) > MV_THRESH_FEAT) next
    y[is.na(y)] <- row_meds[i]
    if (var(y, na.rm = TRUE) == 0) next

    dat <- clin_sub %>% select(all_of(vars_model)) %>% mutate(y = y)
    tryCatch({
      fit <- lm(fml, data = dat); sm <- summary(fit)
      r2_total[i] <- sm$r.squared
      coef_tbl <- sm$coefficients
      for (v in vars_model) {
        if (v %in% rownames(coef_tbl)) {
          coef_mat[i, v] <- coef_tbl[v, "Estimate"]
          pval_mat[i, v] <- coef_tbl[v, "Pr(>|t|)"]
        }
      }
      an <- anova(fit); ss_tot <- sum((y - mean(y))^2)
      for (v in vars_model)
        if (v %in% rownames(an)) r2_mat[i, v] <- an[v, "Sum Sq"] / ss_tot
    }, error = function(e) NULL)
  }

  qval_mat <- apply(pval_mat, 2, p.adjust, method = "BH")
  any_sig  <- rowSums(qval_mat < 0.05, na.rm = TRUE) > 0

  out <- data.frame(mode = mode_name, feature_id = rownames(delta_sub),
                     r2_total_model = r2_total, stringsAsFactors = FALSE)
  for (v in vars_model) {
    out[[paste0("r2_", v)]] <- r2_mat[, v]
    out[[paste0("p_",  v)]] <- pval_mat[, v]
    out[[paste0("q_",  v)]] <- qval_mat[, v]
  }
  out$any_clinical_sig <- any_sig
  out
}

per_feat_results <- lapply(names(delta_all), function(m)
  run_per_feature_lm(delta_all[[m]]$delta,
                     get_clin_df(delta_all[[m]]$patient_ids), m))
names(per_feat_results) <- names(delta_all)
all_feat_df <- bind_rows(per_feat_results)
write.csv(all_feat_df,
          file.path(paths$tables, "04b_clinical_confounder_per_feature.csv"),
          row.names = FALSE)
```

## Confounder summary

```{r conf-summary}
vars_model <- c("age_dx","sex_bin","BMI","stage_bin","location_bin")
var_labels <- c("Age (continuous)","Sex","BMI","Stage","Location")

summary_df <- bind_rows(lapply(seq_along(vars_model), function(j) {
  v <- vars_model[j]; q_col <- paste0("q_", v); r2_col <- paste0("r2_", v)
  total_n <- sum(!is.na(all_feat_df[[q_col]]))
  n_sig   <- sum(all_feat_df[[q_col]] < 0.05, na.rm = TRUE)
  data.frame(variable = var_labels[j], total_tested = total_n,
             n_q05 = n_sig,
             pct_q05 = round(100 * n_sig / max(total_n, 1), 1),
             median_r2 = round(median(all_feat_df[[r2_col]], na.rm = TRUE), 4),
             stringsAsFactors = FALSE)
}))
write.csv(summary_df,
          file.path(paths$tables, "04b_clinical_confounder_summary.csv"),
          row.names = FALSE)
kable(summary_df,
      col.names = c("Variable","Tested","q<0.05","% q<0.05","Median R²"),
      caption = "Clinical effect-modifier screen. Median R² < 0·01 confirms negligible confounding.")
```

## PERMANOVA bubble plot

```{r permanova-plot, fig.height=5}
if (length(permanova_results) > 0) {
  perm_df <- bind_rows(lapply(names(permanova_results), function(m) {
    df <- permanova_results[[m]]; df$mode <- m
    df$variable <- rownames(df); df
  })) %>%
    filter(!variable %in% c("Total","Residual")) %>%
    mutate(variable = recode(variable,
                              age_dx = "Age (continuous)",
                              sex_bin = "Sex", BMI = "BMI",
                              stage_bin = "Stage",
                              location_bin = "Location"),
           sig_label = case_when(`Pr(>F)` < 0.001 ~ "***",
                                  `Pr(>F)` < 0.01  ~ "**",
                                  `Pr(>F)` < 0.05  ~ "*", TRUE ~ "ns"))

  p_perm <- ggplot(perm_df, aes(x = toupper(mode), y = variable,
                                  size = R2, colour = `Pr(>F)` < 0.05)) +
    geom_point(alpha = 0.8) +
    geom_text(aes(label = sig_label), size = 3.5, vjust = -1.0,
              colour = "black") +
    scale_size_continuous(name = "R² (PERMANOVA)", range = c(2, 10)) +
    scale_colour_manual(name = "p < 0.05",
                         values = c("FALSE" = "grey70", "TRUE" = "#C00000"),
                         labels = c("No", "Yes")) +
    labs(title = "PERMANOVA: clinical variables vs. paired-\u0394 metabolome",
         subtitle = "Bubble size = R²; red = p < 0·05 (999 permutations, marginal)",
         x = "LC-MS mode", y = NULL) +
    theme_bw(base_size = 11)
  save_fig(p_perm, "04b_PERMANOVA_bubble", width = 7, height = 5)
  p_perm
}
```

---

# Outputs for downstream analyses

| File | Description |
|------|-------------|
| `output/results/03_statistical_results.rds` | Per-feature Wilcoxon + OPLS-DA — input for `03_biology_translation.Rmd` |
| `output/tables/Table2_differential_metabolites.csv` | Significant features with MSI levels |
| `output/tables/OPLSDA_model_quality.csv` | R²X / R²Y / Q² per mode |
| `output/tables/clinical_metadata_linked.csv` | 182-patient clinical table |
| `output/tables/04b_clinical_*.csv` | Confounder-assessment tables |

---

# Session info

```{r session-info}
sessionInfo()
```
