---
title: "CRC Tissue Metabolomics — Part 3: Biological Interpretation & Translational Analyses"
author: "Chang Gung Memorial Hospital — Metabolomics Medical Research Program"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
params:
  base_dir: "C:/Users/Eric/ClaudeCowork_CRC_metabolomics"
---

```{r base-dir-from-params, include=FALSE}
base_dir <- params$base_dir
```

```{r child = "_setup.Rmd"}
```

```{r extra-libs, include=FALSE}
suppressPackageStartupMessages({
  library(scales)
  library(pheatmap)
  library(ggrepel)
  library(ropls)
  library(pROC)
})
set.seed(42)
```

# Overview

This document performs the biological interpretation and translational
analyses of the statistical results produced by `02_statistics.Rmd`.

| Step | Method | Output |
|------|--------|--------|
| 1. Pathway enrichment | Fisher's exact ORA over 19 curated CRC pathways | Table 3, Figure 6 |
| 2. Colon sidedness | Right vs. Left tumour (within-tumour + paired FC) | Figures S1–S4 |
| 3. ROC biomarker panel | Composite scoring + 10-fold CV logistic regression | Figure 8, biomarker table |

**Inputs:**
- `output/tables/Table2_differential_metabolites.csv`
- `output/results/03_statistical_results.rds`
- `output/tables/clinical_metadata_linked.csv`
- `output/results/02b_batch_corrected_datasets.rds`

**Expected render time: 5–10 min** — fastest Rmd; consumes pre-computed
`.rds` files and runs Fisher tests, small OPLS-DA, and CV logistic regression.

---

# 1. Pathway Enrichment (ORA)

## Load significant features and full universe

```{r pathway-load}
sig_df <- read.csv(file.path(paths$tables, "Table2_differential_metabolites.csv"),
                   stringsAsFactors = FALSE)
stat   <- readRDS(file.path(paths$results, "03_statistical_results.rds"))
all_df <- stat$all_results

clean_name <- function(x) str_replace(x, "^(low score|no MS2): ", "")
all_df$clean_name <- clean_name(all_df$metabolite)
sig_df$clean_name <- clean_name(sig_df$metabolite)

all_named <- all_df %>% filter(metabolite != "Unknown", msi_level %in% c(2, 3))
sig_named <- sig_df %>% filter(metabolite != "Unknown", msi_level %in% c(2, 3))

cat(sprintf("Universe (named, MSI 2-3): %d (unique names: %d)\n",
            nrow(all_named), n_distinct(all_named$clean_name)))
cat(sprintf("Significant (named, MSI 2-3): %d (unique names: %d)\n",
            nrow(sig_named), n_distinct(sig_named$clean_name)))
```

## Curated pathway database

Pathways defined by **metabolite names** (central carbon, amino acids, bile
acids, etc.) and **lipid ontology classes** from LipidBlast (sphingolipids,
phospholipids, glycerolipids, fatty acids).

```{r pathway-db}
pathway_db <- list(
  list(id="hsa00010", name="Glycolysis / Gluconeogenesis",
       members=c("Glucose","Glucose-1-phosphate","Glucose-6-phosphate",
                  "Fructose-6-phosphate","Fructose-1,6-bisphosphate",
                  "Glyceraldehyde-3-phosphate","Pyruvate","Lactate",
                  "Lactic acid","Phosphoenolpyruvate"),
       category="Central carbon"),
  list(id="hsa00020", name="TCA cycle",
       members=c("Citric acid","Citrate","Isocitrate","alpha-Ketoglutarate",
                  "Succinate","Succinic acid","Fumarate","Fumaric acid",
                  "Malate","Malic acid","Oxaloacetate",
                  "Glutamic acid","Glutamate","Aspartic acid","Aspartate"),
       category="Central carbon"),
  list(id="hsa00480", name="Glutathione metabolism",
       members=c("Glutathione","GSSG","GSH","gamma-Glutamylcysteine",
                  "gamma-Glutamylleucine","gamma-Glu-Tyr",
                  "Glutamic acid","Glutamate","Cystine","Cysteine",
                  "Glycine","Pyroglutamic acid","5-Oxoproline","Ornithine"),
       category="Redox"),
  list(id="hsa00230", name="Purine metabolism",
       members=c("Adenine","Guanine","Hypoxanthine","Xanthine",
                  "Adenosine","Guanosine","Inosine","Xanthosine",
                  "IMP","GMP","AMP","S-Adenosyl-homocysteine",
                  "2'-Deoxyguanosine","2'-Deoxyadenosine"),
       category="Nucleotide biosynthesis"),
  list(id="hsa00240", name="Pyrimidine metabolism",
       members=c("Uridine","Cytidine","Thymidine","UMP","CMP",
                  "3-Methylcytidine","Uracil","Thymine","Cytosine"),
       category="Nucleotide biosynthesis"),
  list(id="hsa00270", name="One-carbon / Methionine cycle",
       members=c("S-Adenosyl-homocysteine","S-Adenosylmethionine",
                  "Methionine","Homocysteine","Homoserine",
                  "Serine","Glycine","Betaine",
                  "Folate","5-Methyltetrahydrofolate","Cystathionine"),
       category="One-carbon"),
  list(id="hsa00380", name="Tryptophan / Kynurenine pathway",
       members=c("Tryptophan","Kynurenine","Kynurenic acid",
                  "3-Hydroxykynurenine","Quinolinic acid",
                  "Serotonin","5-Hydroxyindoleacetic acid",
                  "Indole-3-acetic acid","Indole","Indole-3-carboxaldehyde",
                  "Melatonin","Nicotinamide","NAD","Nicotinic acid"),
       category="Amino acid"),
  list(id="hsa00330", name="Arginine / Polyamine metabolism",
       members=c("Arginine","Ornithine","Citrulline","Proline",
                  "Putrescine","Spermidine","Spermine","N-Acetylputrescine",
                  "Glutamic acid","Glutamate","Creatine","Creatinine",
                  "4-Aminobutyric acid","GABA","Glutamine"),
       category="Amino acid / Polyamine"),
  list(id="hsa00340", name="Histidine metabolism",
       members=c("Histidine","Histamine","Imidazole-4-acetaldehyde",
                  "Urocanate","Carnosine","Anserine",
                  "1-Methylhistidine","3-Methylhistidine"),
       category="Amino acid"),
  list(id="hsa00280", name="BCAA metabolism",
       members=c("Valine","Leucine","Isoleucine",
                  "gamma-Glutamylleucine","Ala-Val","Ala-Ile","Gly-Val"),
       category="Amino acid"),
  list(id="hsa00120", name="Bile acid metabolism",
       members=c("Cholic acid","Chenodeoxycholic acid",
                  "Deoxycholic acid","Lithocholic acid",
                  "Ursodeoxycholic acid","Taurine","Taurocholic acid",
                  "Taurodeoxycholic acid","Taurochenodeoxycholic acid",
                  "Glycocholic acid","Glycodeoxycholic acid",
                  "Glycochenodeoxycholic acid"),
       category="Bile acid"),
  list(id="hsa00430", name="Taurine / hypotaurine metabolism",
       members=c("Taurine","Hypotaurine","Cysteic acid",
                  "Cysteine sulfinic acid"),
       category="Amino acid"),
  list(id="hsa00270b", name="Cysteine / Methionine metabolism",
       members=c("Methionine","Cystine","Cysteine","S-Adenosyl-homocysteine",
                  "Homoserine","Taurine","Pyroglutamic acid"),
       category="Amino acid"),
  list(id="hsa00310", name="Lysine degradation",
       members=c("Lysine","N-epsilon-Acetyllysine","Pipecolic acid",
                  "Cadaverine","Saccharopine","2-Aminoadipic acid"),
       category="Amino acid"),
  list(id="hsa00590", name="Arachidonic acid / Eicosanoids",
       members=c("Arachidonic acid","Prostaglandin E1","Prostaglandin E2",
                  "Prostaglandin D2","Prostaglandin F2alpha",
                  "Thromboxane B2","Leukotriene B4",
                  "HETE","5-HETE","12-HETE","15-HETE","FA 20:4"),
       category="Lipid signalling"),
  list(id="hsa00600", name="Sphingolipid metabolism",
       members=NULL,
       ontology=c("Cer_NS","Cer_NDS","Cer_ADS","Cer_BDS",
                   "SM","SHexCer","HexCer_AP","Hex2Cer",
                   "Cer_AP","Cer_AS","HexCer_NS","HexCer_NDS"),
       category="Lipid"),
  list(id="hsa00564", name="Glycerophospholipid metabolism",
       members=c("O-Phosphocolamine"),
       ontology=c("PC","PE","PS","PI","PA","PG",
                   "LPC","LPE","LPS","LPI","LPG","LPA",
                   "EtherPC","EtherPE","EtherLPC","EtherLPE",
                   "EtherLPG","LNAPE"),
       category="Lipid"),
  list(id="hsa00561", name="Glycerolipid metabolism",
       members=NULL,
       ontology=c("DG","TG","MG","EtherDG","EtherTG","EtherMG"),
       category="Lipid"),
  list(id="hsa01212", name="Fatty acid metabolism",
       members=c("Linoleic acid","Oleic acid","Palmitic acid",
                  "Stearic acid","Abietic acid"),
       ontology=c("FA","FAHFA","DMEDFAHFA","CAR",
                   "Lineolic acids and derivatives",
                   "Long-chain fatty acids",
                   "Fatty acid esters","Fatty acid methyl esters"),
       category="Lipid")
)
cat(sprintf("Pathway database: %d pathways\n", length(pathway_db)))
```

## Fisher's exact test per pathway

```{r pathway-ora}
find_pathway_features <- function(pw, data) {
  idx <- rep(FALSE, nrow(data))
  if (!is.null(pw$members))
    idx <- idx | (tolower(data$clean_name) %in% tolower(pw$members))
  if (!is.null(pw$ontology))
    idx <- idx | (data$ontology %in% pw$ontology)
  idx
}

n_universe <- nrow(all_named)
n_sig      <- nrow(sig_named)

# Cache membership per pathway
pathway_sig_cache <- setNames(
  lapply(pathway_db, function(pw) find_pathway_features(pw, sig_named)),
  sapply(pathway_db, function(pw) pw$id))

enrichment_results <- data.frame()
for (pw in pathway_db) {
  in_univ <- find_pathway_features(pw, all_named)
  in_sig  <- pathway_sig_cache[[pw$id]]
  n_pw_univ <- sum(in_univ); n_pw_sig <- sum(in_sig)

  if (n_pw_sig == 0 || n_pw_univ == 0) next

  a <- n_pw_sig; b <- n_sig - n_pw_sig
  c_val <- max(0, n_pw_univ - n_pw_sig)
  d     <- max(0, (n_universe - n_sig) - c_val)

  ft <- fisher.test(matrix(c(a, c_val, b, d), nrow = 2),
                    alternative = "greater")
  expected <- n_pw_univ * (n_sig / n_universe)
  er       <- a / max(expected, 0.01)
  hit_names <- sort(unique(sig_named$clean_name[in_sig]))

  enrichment_results <- rbind(enrichment_results, data.frame(
    pathway_id       = pw$id,
    pathway_name     = pw$name,
    category         = pw$category,
    hits_sig         = a,
    hits_universe    = n_pw_univ,
    enrichment_ratio = round(er, 3),
    p_value          = ft$p.value,
    hit_metabolites  = paste(hit_names, collapse = "; "),
    stringsAsFactors = FALSE))
}
enrichment_results$q_value <- p.adjust(enrichment_results$p_value, method = "BH")
enrichment_results <- enrichment_results %>% arrange(p_value)
```

## Direction summary per pathway

```{r direction-summary}
dir_summary <- bind_rows(lapply(seq_len(nrow(enrichment_results)), function(i) {
  r      <- enrichment_results[i, ]
  in_sig <- pathway_sig_cache[[r$pathway_id]]
  hits   <- sig_named[in_sig, ] %>%
    group_by(direction) %>%
    summarise(n_unique = n_distinct(clean_name),
              mean_fc  = mean(fc, na.rm = TRUE),
              .groups  = "drop")

  n_up      <- sum(hits$n_unique[hits$direction == "Up_in_Tumour"])
  n_down    <- sum(hits$n_unique[hits$direction == "Down_in_Tumour"])
  fc_up     <- if (any(hits$direction == "Up_in_Tumour"))
                  hits$mean_fc[hits$direction == "Up_in_Tumour"] else NA
  fc_down   <- if (any(hits$direction == "Down_in_Tumour"))
                  hits$mean_fc[hits$direction == "Down_in_Tumour"] else NA

  data.frame(pathway_id = r$pathway_id, pathway_name = r$pathway_name,
             up_unique = n_up, down_unique = n_down,
             mean_fc_up = round(fc_up, 2), mean_fc_down = round(fc_down, 2),
             p_value = r$p_value, q_value = r$q_value,
             stringsAsFactors = FALSE)
}))
```

## Table 3 — enriched pathways

```{r table3}
table3 <- enrichment_results %>%
  left_join(dir_summary %>% select(pathway_id, up_unique, down_unique,
                                    mean_fc_up, mean_fc_down),
            by = "pathway_id") %>%
  mutate(p_value = signif(p_value, 3),
         q_value = signif(q_value, 3),
         significance = case_when(q_value < 0.001 ~ "***",
                                   q_value < 0.01  ~ "**",
                                   q_value < 0.05  ~ "*",
                                   q_value < 0.1   ~ ".",
                                   TRUE            ~ "ns"))
write.csv(table3, file.path(paths$tables, "Table3_enriched_pathways.csv"),
          row.names = FALSE)
kable(table3 %>% select(pathway_name, category, hits_sig, hits_universe,
                          enrichment_ratio, p_value, q_value,
                          up_unique, down_unique, significance),
      col.names = c("Pathway","Category","Hits (sig)","Hits (univ)","ER","p","q",
                     "Up","Down","Sig"),
      caption = "Table 3 — pathway enrichment (Fisher exact, BH-corrected).")
```

## Figure 6 — enrichment bubble plot

```{r bubble-plot, fig.height=7}
plot_df <- enrichment_results %>%
  mutate(neg_log10_p   = -log10(p_value),
         pathway_label = paste0(pathway_name, " (", hits_sig, "/",
                                 hits_universe, ")")) %>%
  arrange(enrichment_ratio) %>%
  mutate(pathway_label = factor(pathway_label, levels = pathway_label))

p_bubble <- ggplot(plot_df, aes(x = enrichment_ratio, y = pathway_label)) +
  geom_point(aes(size = hits_sig, colour = neg_log10_p), alpha = 0.85) +
  scale_colour_gradient(low = "#FDE725", high = "#440154",
                         name = expression(-log[10](italic(p)))) +
  scale_size_continuous(range = c(3, 14), name = "Sig. hits") +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  labs(x = "Enrichment Ratio (observed / expected)", y = NULL,
       title = "Metabolic pathway enrichment — CRC tumour vs. non-tumour",
       subtitle = sprintf("ORA (Fisher exact, BH) | %d sig / %d measured | %d pathways FDR<0.05",
                           n_sig, n_universe,
                           sum(enrichment_results$q_value < 0.05))) +
  theme_bw(base_size = 12)
if (any(enrichment_results$q_value < 0.05)) {
  sig_pts <- plot_df %>% filter(q_value < 0.05)
  p_bubble <- p_bubble + geom_point(data = sig_pts, shape = 1,
                                      colour = "red", stroke = 1.0, size = 5)
}
save_fig(p_bubble, "06_Pathway_enrichment_bubble", width = 12, height = 7)
p_bubble
```

## Figure 6b — direction bar chart

```{r direction-bar, fig.height=6}
dir_plot_df <- dir_summary %>%
  pivot_longer(cols = c(up_unique, down_unique),
               names_to = "direction", values_to = "count") %>%
  mutate(direction = ifelse(direction == "up_unique",
                             "Up in Tumour", "Down in Tumour")) %>%
  filter(count > 0)
pw_order <- dir_summary %>%
  mutate(total = up_unique + down_unique) %>%
  arrange(total) %>% pull(pathway_name)
dir_plot_df$pathway_name <- factor(dir_plot_df$pathway_name, levels = pw_order)

p_dir <- ggplot(dir_plot_df,
                 aes(x = count, y = pathway_name, fill = direction)) +
  geom_col(position = "stack", width = 0.6, alpha = 0.85) +
  scale_fill_manual(values = c("Up in Tumour" = "#D73027",
                                "Down in Tumour" = "#4575B4"),
                    name = "Direction") +
  labs(x = "Number of significant metabolites (unique)", y = NULL,
       title = "Direction of metabolite changes per pathway") +
  theme_bw(base_size = 12)
save_fig(p_dir, "06b_Pathway_direction_bar", width = 10, height = 6.5)
p_dir
```

---

# 2. Colon Sidedness Analysis

Right-sided (proximal, midgut) vs. left-sided (distal, hindgut) colon
cancers differ molecularly and prognostically. This section tests whether
the tumour metabolome differs by side in the colon-only subset.

## Load data and define sides

```{r sidedness-load}
dat  <- readRDS(file.path(paths$results,
                           "02b_batch_corrected_datasets.rds"))
clin <- read.csv(file.path(paths$tables, "clinical_metadata_linked.csv"),
                 stringsAsFactors = FALSE)
clin$core_id <- str_replace(clin$serial_id, "CO", "C0")

mode_names  <- c("meta_pos","meta_neg","lip_pos","lip_neg")
mode_labels <- c("Metabolomics ESI+","Metabolomics ESI-",
                 "Lipidomics ESI+",   "Lipidomics ESI-")

colon_clin <- clin[clin$location == "Colon" &
                     clin$colon_side %in% c("Right","Left"), ]
cat(sprintf("Colon patients: %d (Right=%d, Left=%d)\n",
            nrow(colon_clin),
            sum(colon_clin$colon_side == "Right"),
            sum(colon_clin$colon_side == "Left")))

get_side_map <- function(pair_info) {
  m <- match(pair_info$patient_base, colon_clin$core_id)
  keep <- !is.na(m)
  side <- colon_clin$colon_side[m[keep]]
  names(side) <- pair_info$patient_id[keep]
  side
}
```

## Within-tumour Wilcoxon (Right vs Left)

```{r sidedness-stats, cache=TRUE}
all_results <- list()
for (i in seq_along(mode_names)) {
  mn <- mode_names[i]; ml <- mode_labels[i]
  cat(sprintf("\n%s ... ", ml))

  p  <- dat[[mn]]$paired
  fm <- dat[[mn]]$feature_meta
  side_map <- get_side_map(p$pair_info)
  common_R <- intersect(names(side_map)[side_map == "Right"],
                         colnames(p$area_T))
  common_L <- intersect(names(side_map)[side_map == "Left"],
                         colnames(p$area_T))
  n_features <- nrow(p$area_T)

  # Tumour Right vs Left (Wilcoxon rank-sum)
  tumour_pval <- tumour_fc <- rep(NA_real_, n_features)
  for (j in seq_len(n_features)) {
    vR <- as.numeric(p$area_T[j, common_R]); vR <- vR[!is.na(vR)]
    vL <- as.numeric(p$area_T[j, common_L]); vL <- vL[!is.na(vL)]
    if (length(vR) >= 3 && length(vL) >= 3) {
      wt <- suppressWarnings(wilcox.test(vR, vL))
      tumour_pval[j] <- wt$p.value
      tumour_fc[j]   <- mean(vR, na.rm=TRUE) / mean(vL, na.rm=TRUE)
    }
  }
  tumour_qval <- p.adjust(tumour_pval, method = "BH")

  # Paired FC shift: log2(T/N)_R vs log2(T/N)_L
  paired_pval <- median_R <- median_L <- rep(NA_real_, n_features)
  all_colon <- c(common_R, common_L)
  for (j in seq_len(n_features)) {
    logfc <- log2((as.numeric(p$area_T[j, all_colon]) + 1) /
                  (as.numeric(p$area_N[j, all_colon]) + 1))
    names(logfc) <- all_colon
    lR <- logfc[common_R]; lR <- lR[!is.na(lR) & is.finite(lR)]
    lL <- logfc[common_L]; lL <- lL[!is.na(lL) & is.finite(lL)]
    if (length(lR) >= 3 && length(lL) >= 3) {
      wt <- suppressWarnings(wilcox.test(lR, lL))
      paired_pval[j] <- wt$p.value
      median_R[j] <- median(lR); median_L[j] <- median(lL)
    }
  }
  paired_qval <- p.adjust(paired_pval, method = "BH")

  all_results[[mn]] <- list(
    res = data.frame(
      feature_id = fm$alignment_id, metabolite = fm$metabolite,
      mz = fm$mz, rt_min = fm$rt_min, adduct = fm$adduct,
      ontology = fm$ontology, mode = mn, mode_label = ml,
      tumour_pval = tumour_pval, tumour_qval = tumour_qval,
      tumour_fc_RL = tumour_fc, log2fc_RL = log2(tumour_fc),
      paired_fc_pval = paired_pval, paired_fc_qval = paired_qval,
      median_logfc_R = median_R, median_logfc_L = median_L,
      delta_logfc = median_R - median_L,
      stringsAsFactors = FALSE),
    area_T = p$area_T, area_N = p$area_N,
    common_R = common_R, common_L = common_L, fm = fm)

  cat(sprintf("sig tumour R vs L (q<0.05): %d / %d\n",
              sum(tumour_qval < 0.05, na.rm = TRUE), n_features))
}

all_res_df <- do.call(rbind, lapply(all_results, function(x) x$res))
write.csv(all_res_df[order(all_res_df$tumour_pval), ],
          file.path(paths$tables, "Table_sidedness_differential.csv"),
          row.names = FALSE)
```

## Volcano plots — Right vs Left tumour

```{r sidedness-volcano, fig.height=10}
volcano_list <- list()
for (i in seq_along(mode_names)) {
  mn <- mode_names[i]; ml <- mode_labels[i]
  res <- all_results[[mn]]$res
  res$sig <- ifelse(!is.na(res$tumour_qval) & res$tumour_qval < 0.05,
                     "Significant", "NS")
  res$label <- ifelse(res$sig == "Significant" & res$metabolite != "Unknown",
                       res$metabolite, "")
  res$label <- ifelse(nchar(res$label) > 35,
                       paste0(substr(res$label, 1, 32), "..."), res$label)
  res$nlog10q <- pmin(-log10(res$tumour_qval), 15, na.rm = TRUE)

  volcano_list[[i]] <- ggplot(res, aes(x = log2fc_RL, y = nlog10q, color = sig)) +
    geom_point(alpha = 0.5, size = 1) +
    scale_color_manual(values = c("Significant" = "red3", "NS" = "grey60")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "blue") +
    geom_vline(xintercept = c(-1, 1), linetype = "dotted", color = "grey40") +
    geom_text_repel(aes(label = label), size = 2, max.overlaps = 12,
                    color = "black", segment.color = "grey50") +
    labs(title = ml, x = expression(log[2]~"FC (R / L)"),
         y = expression(-log[10]~"q")) +
    theme_bw(base_size = 10) + theme(legend.position = "none")
}
fig_s1 <- wrap_plots(volcano_list, ncol = 2) +
  plot_annotation(title = "Right vs. Left colon — tumour metabolite differences")
save_fig(fig_s1, "FigS1_sidedness_volcano", width = 12, height = 10)
fig_s1
```

## OPLS-DA — best mode

```{r sidedness-opls, cache=TRUE, fig.height=6}
sig_counts <- sapply(all_results,
                      function(x) sum(x$res$tumour_qval < 0.05, na.rm = TRUE))
best_mode <- if (max(sig_counts) > 0) names(which.max(sig_counts)) else
              names(which.min(sapply(all_results,
                function(x) median(x$res$tumour_pval, na.rm = TRUE))))
cat(sprintf("Best mode for OPLS-DA: %s (%d sig)\n", best_mode, sig_counts[best_mode]))

ar   <- all_results[[best_mode]]
ids  <- c(ar$common_R, ar$common_L)
mat  <- t(ar$area_T[, ids])
side <- factor(ifelse(rownames(mat) %in% ar$common_R, "Right", "Left"),
                levels = c("Left","Right"))

col_var <- apply(mat, 2, var, na.rm = TRUE)
keep    <- which(!is.na(col_var) & col_var > 0)
mat     <- mat[, keep]
mat     <- impute_row_median(t(mat)); mat <- t(mat)
mat_sc  <- scale(log2(mat + 1))

opls_sd <- tryCatch(
  opls(mat_sc, side, predI = 1, orthoI = NA, permI = 1000,
       fig.pdfC = "none", info.txtC = "none"),
  error = function(e) tryCatch(
    opls(mat_sc, side, predI = 1, orthoI = 0, permI = 1000,
         fig.pdfC = "none", info.txtC = "none"),
    error = function(e2) NULL))

if (!is.null(opls_sd)) {
  r2y <- opls_sd@summaryDF$`R2Y(cum)`
  q2  <- opls_sd@summaryDF$`Q2(cum)`
  pQ2 <- opls_sd@summaryDF$pQ2
  cat(sprintf("  R²Y=%.3f  Q²=%.3f  pQ²=%.3f\n", r2y, q2, pQ2))

  scores <- data.frame(
    t1   = getScoreMN(opls_sd)[, 1],
    Side = side)
  has_ortho <- ncol(getScoreMN(opls_sd, orthoL = TRUE)) > 0
  scores$to1 <- if (has_ortho) getScoreMN(opls_sd, orthoL = TRUE)[, 1]
                else getScoreMN(opls_sd)[, 2]

  fig_s2 <- ggplot(scores, aes(x = t1, y = to1, color = Side, fill = Side)) +
    geom_point(size = 2.5, alpha = 0.7) +
    stat_ellipse(level = 0.95, alpha = 0.1, geom = "polygon") +
    scale_color_manual(values = c("Right" = "#E74C3C", "Left" = "#3498DB")) +
    scale_fill_manual(values = c("Right" = "#E74C3C", "Left" = "#3498DB")) +
    labs(title = sprintf("OPLS-DA — %s tumours (R vs L)",
                          mode_labels[match(best_mode, mode_names)]),
         subtitle = sprintf("R²Y=%.3f  Q²=%.3f  pQ²=%.3f", r2y, q2, pQ2),
         x = "t_pred[1]",
         y = if (has_ortho) "t_orth[1]" else "t[2]") +
    theme_bw(base_size = 12)
  save_fig(fig_s2, "FigS2_sidedness_OPLSDA", width = 8, height = 6)
  fig_s2
}
```

## Sidedness outputs summary

```{r sidedness-summary}
ss_summary <- data.frame(
  Mode       = mode_labels,
  Sig_TvsN   = sapply(all_results,
                       function(x) sum(x$res$tumour_qval < 0.05, na.rm = TRUE)),
  Sig_paired = sapply(all_results,
                       function(x) sum(x$res$paired_fc_qval < 0.05, na.rm = TRUE)),
  Total      = sapply(all_results, function(x) nrow(x$res)))
kable(ss_summary,
      col.names = c("Mode","R vs L tumour (q<0.05)",
                     "T/N shift (q<0.05)","Total features"),
      caption = "Sidedness analysis — significant features per mode.")

saveRDS(list(all_results_df = all_res_df, colon_clinical = colon_clin,
              oplsda_model = if (exists("opls_sd")) opls_sd else NULL,
              oplsda_mode = best_mode),
        file.path(paths$results, "06_sidedness_results.rds"))
```

---

# 3. ROC Biomarker Panel

## Candidate selection

Criteria:

- q < 0·01 (stringent FDR)
- OPLS-DA VIP ≥ 2·0 (robust multivariate importance)
- MSI confidence level 2 or 3 (named, interpretable)
- |log2FC| ≥ 1 (≥ 2-fold)

```{r roc-candidates}
vip_all <- do.call(rbind, lapply(names(stat$opls_results), function(m) {
  v <- stat$opls_results[[m]]$vip
  if (is.null(v)) return(NULL)
  v$mode <- m; v
}))
ar <- merge(stat$all_results, vip_all[, c("alignment_id","mode","vip")],
            by = c("alignment_id","mode"), all.x = TRUE)

candidates <- ar[!is.na(ar$q_value) & !is.na(ar$vip) &
                   ar$q_value < 0.01 & ar$vip >= 2.0 &
                   ar$msi_level %in% c(2, 3) &
                   abs(ar$log2fc) >= 1, ]
candidates$composite <- -log10(candidates$q_value) *
                         candidates$vip * abs(candidates$log2fc)
candidates <- candidates[order(-candidates$composite), ]

n_top <- min(20, nrow(candidates))
top20 <- candidates[1:n_top, ]
cat(sprintf("Candidates passing all criteria: %d (using top %d)\n",
            nrow(candidates), n_top))
```

## Individual ROC curves

```{r roc-individual}
bc <- dat   # already loaded above

build_roc_data <- function(feature_row, bc_data) {
  mode <- feature_row$mode; aid <- as.character(feature_row$alignment_id)
  paired <- bc_data[[mode]]$paired
  if (!aid %in% rownames(paired$area_T)) return(NULL)
  vT <- as.numeric(paired$area_T[aid, ])
  vN <- as.numeric(paired$area_N[aid, ])
  data.frame(value = c(vT, vN),
             class = rep(c(1L, 0L), each = length(vT)),
             patient = rep(colnames(paired$area_T), 2),
             stringsAsFactors = FALSE)
}

roc_data_list <- lapply(seq_len(n_top), function(i) {
  rd <- build_roc_data(top20[i, ], bc)
  if (is.null(rd)) return(NULL)
  rd[complete.cases(rd), ]
})
valid <- !sapply(roc_data_list, is.null)

roc_results <- vector("list", n_top)
for (i in seq_len(n_top)) {
  if (!valid[i]) next
  rd <- roc_data_list[[i]]
  ro <- roc(rd$class, rd$value, quiet = TRUE, direction = "auto")
  cb <- coords(ro, "best",
               ret = c("threshold","sensitivity","specificity"),
               best.method = "youden")
  roc_results[[i]] <- list(roc = ro, auc = as.numeric(auc(ro)),
                           sens = cb$sensitivity[1], spec = cb$specificity[1])
}
```

## Multi-marker panel with 10-fold CV

```{r roc-panel, cache=TRUE}
build_panel_matrix <- function(top_n, top20_df, bc_data) {
  rd_list <- lapply(seq_len(top_n), function(i) {
    rd <- build_roc_data(top20_df[i, ], bc_data)
    if (!is.null(rd)) rd[complete.cases(rd), ] else NULL
  })
  patient_sets <- lapply(rd_list,
                          function(rd) if (is.null(rd)) character(0)
                                        else unique(rd$patient))
  common_patients <- Reduce(intersect,
                             patient_sets[!sapply(patient_sets,
                                                    function(x) length(x)==0)])
  mat_cols <- lapply(rd_list, function(rd) {
    if (is.null(rd)) return(NULL)
    s <- rd[rd$patient %in% common_patients, ]
    s <- s[order(-s$class, s$patient), ]
    s$value
  })
  rd_ref <- rd_list[[which(!sapply(rd_list, is.null))[1]]]
  rd_ref <- rd_ref[rd_ref$patient %in% common_patients, ]
  rd_ref <- rd_ref[order(-rd_ref$class, rd_ref$patient), ]

  mat <- do.call(cbind, mat_cols); colnames(mat) <- paste0("M", seq_len(ncol(mat)))
  data.frame(class = rd_ref$class, patient = rd_ref$patient, mat,
             stringsAsFactors = FALSE)
}

cv_roc_panel <- function(panel_df, n_folds = 10) {
  patients <- unique(panel_df$patient)
  fold_assign <- sample(rep(seq_len(n_folds), length.out = length(patients)))
  patient_fold <- setNames(fold_assign, patients)
  panel_df$fold <- patient_fold[as.character(panel_df$patient)]

  pred_all  <- numeric(nrow(panel_df))
  feat_cols <- grep("^M", names(panel_df), value = TRUE)
  fmla <- as.formula(paste("class ~", paste(feat_cols, collapse = " + ")))

  for (f in seq_len(n_folds)) {
    train <- panel_df[panel_df$fold != f, ]
    test  <- panel_df[panel_df$fold == f, ]
    fit <- suppressWarnings(glm(fmla, data = train,
                                 family = binomial,
                                 control = glm.control(maxit = 500)))
    pred_all[panel_df$fold == f] <- predict(fit, newdata = test,
                                             type = "response")
  }
  ro <- roc(panel_df$class, pred_all, quiet = TRUE)
  list(roc = ro, auc = as.numeric(auc(ro)))
}

panel_sizes   <- c(5, 10, 20)
panel_results <- list()
for (ps in panel_sizes) {
  pmat <- build_panel_matrix(ps, top20, bc)
  aucs <- replicate(5, cv_roc_panel(pmat)$auc)
  best_res <- cv_roc_panel(pmat)
  panel_results[[as.character(ps)]] <- list(
    roc = best_res$roc, auc = best_res$auc,
    auc_mean = mean(aucs), auc_sd = sd(aucs), n_markers = ps)
  cat(sprintf("  %d-marker panel: CV-AUC = %.3f (mean=%.3f, sd=%.3f)\n",
              ps, best_res$auc, mean(aucs), sd(aucs)))
}

best_ind_idx <- which.max(sapply(roc_results,
                                    function(x) if (is.null(x)) 0 else x$auc))
best_ind_auc <- roc_results[[best_ind_idx]]$auc
cat(sprintf("\nBest individual: %s (AUC=%.3f)\n",
            top20$metabolite[best_ind_idx], best_ind_auc))
```

## Figure 8 — ROC panels

```{r figure-8, fig.height=12}
panel_cols <- c("#E41A1C","#377EB8","#4DAF4A","#984EA3","#FF7F00")

# Fig 8A: top 5 individual curves
auc_vals <- sapply(seq_len(n_top),
                    function(i) if (is.null(roc_results[[i]])) 0
                                else roc_results[[i]]$auc)
top5_idx <- order(-auc_vals)[1:5]
fig8a_data <- do.call(rbind, lapply(seq_along(top5_idx), function(j) {
  i <- top5_idx[j]; r <- roc_results[[i]]$roc
  nm <- gsub("low score: ", "", top20$metabolite[i])
  if (nchar(nm) > 35) nm <- paste0(substr(nm, 1, 32), "...")
  data.frame(sens = r$sensitivities, spec = 1 - r$specificities,
             metabolite = sprintf("%s (AUC=%.3f)", nm,
                                   roc_results[[i]]$auc),
             stringsAsFactors = FALSE)
}))
fig8a <- ggplot(fig8a_data, aes(x = spec, y = sens, color = metabolite)) +
  geom_line(linewidth = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = panel_cols) +
  labs(title = "A", x = "1 - Specificity", y = "Sensitivity", color = NULL) +
  theme_bw(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 14),
        legend.position = c(0.65, 0.2),
        legend.text = element_text(size = 7)) +
  coord_equal()

# Fig 8B: panel ROCs + best individual
fig8b_data <- do.call(rbind, lapply(as.character(panel_sizes), function(ps) {
  r <- panel_results[[ps]]$roc
  data.frame(sens = r$sensitivities, spec = 1 - r$specificities,
             panel = sprintf("%s-marker panel (AUC=%.3f)",
                              ps, panel_results[[ps]]$auc),
             stringsAsFactors = FALSE)
}))
best_r <- roc_results[[best_ind_idx]]$roc
best_nm <- gsub("low score: ", "", top20$metabolite[best_ind_idx])
fig8b_data <- rbind(fig8b_data, data.frame(
  sens = best_r$sensitivities, spec = 1 - best_r$specificities,
  panel = sprintf("Best individual: %s (AUC=%.3f)", best_nm, best_ind_auc),
  stringsAsFactors = FALSE))
fig8b <- ggplot(fig8b_data, aes(x = spec, y = sens, color = panel)) +
  geom_line(linewidth = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c(panel_cols[1:3], "grey40")) +
  labs(title = "B", x = "1 - Specificity", y = "Sensitivity", color = NULL) +
  theme_bw(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 14),
        legend.position = c(0.62, 0.2),
        legend.text = element_text(size = 7)) +
  coord_equal()

# Fig 8C: AUC bar plot for top 20
auc_bar <- data.frame(
  metabolite = sapply(seq_len(n_top), function(i) {
    nm <- gsub("low score: ", "", top20$metabolite[i])
    if (nchar(nm) > 30) nm <- paste0(substr(nm, 1, 27), "...")
    nm}),
  auc = sapply(seq_len(n_top),
                function(i) if (is.null(roc_results[[i]])) NA
                              else roc_results[[i]]$auc),
  direction = top20$direction,
  stringsAsFactors = FALSE)
auc_bar$metabolite <- factor(auc_bar$metabolite, levels = rev(auc_bar$metabolite))

fig8c <- ggplot(auc_bar, aes(x = metabolite, y = auc, fill = direction)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = c(0.7, 0.8), linetype = c("dashed","dotted"),
             color = "grey40") +
  geom_text(aes(label = sprintf("%.3f", auc)), hjust = -0.1, size = 2.8) +
  scale_fill_manual(values = c("Up_in_Tumour" = "#E41A1C",
                                "Down_in_Tumour" = "#377EB8"),
                    labels = c("Up_in_Tumour" = "Elevated",
                                "Down_in_Tumour" = "Decreased")) +
  scale_y_continuous(limits = c(0, 1.05), breaks = seq(0, 1, 0.1)) +
  coord_flip() +
  labs(title = "C", x = NULL, y = "AUC", fill = "In tumour") +
  theme_bw(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 14),
        legend.position = "bottom")

fig8 <- (fig8a | fig8b) / fig8c + plot_layout(heights = c(1, 1.2)) +
  plot_annotation(title = "Figure 8 — Diagnostic biomarker panel (ROC analysis)",
                  subtitle = "CRC tumour vs. adjacent normal tissue (paired n = 182)")
save_fig(fig8, "Figure_08_ROC_biomarker_panel", width = 14, height = 12)
fig8
```

## Biomarker panel table

```{r biomarker-table}
tbl <- data.frame(
  Rank       = seq_len(n_top),
  Metabolite = top20$metabolite,
  Mode       = top20$mode,
  m_z        = round(top20$mz, 4),
  RT_min     = round(top20$rt_min, 2),
  MSI        = top20$msi_level,
  Direction  = top20$direction,
  Fold_Change = round(2^abs(top20$log2fc), 2),
  log2FC     = round(top20$log2fc, 3),
  q_value    = signif(top20$q_value, 3),
  VIP        = round(top20$vip, 3),
  Composite  = round(top20$composite, 1),
  AUC        = round(sapply(seq_len(n_top),
                     function(i) if (is.null(roc_results[[i]])) NA
                                   else roc_results[[i]]$auc), 3))
write.csv(tbl, file.path(paths$tables, "Table_biomarker_panel.csv"),
          row.names = FALSE)
kable(tbl, caption = "Top 20 biomarker candidates with ROC AUC.")

saveRDS(list(top20_candidates = top20, roc_results = roc_results,
              panel_results = panel_results, biomarker_table = tbl,
              best_individual = list(metabolite = top20$metabolite[best_ind_idx],
                                      auc = best_ind_auc)),
        file.path(paths$results, "07_ROC_results.rds"))
```

## Translational outlook

Candidate metabolites here constitute a **tissue-level discovery panel**.
Before clinical use they require:

1. External validation in an independent CRC tissue cohort
2. Cross-sample-type validation in plasma, stool, and/or urine
3. Multi-centre confirmation (this single-centre East Asian cohort limits
   current generalisability)

Tissue signals with known presence in accessible biofluids (e.g., taurine,
specific LPCs) are priority candidates for the **Year 3–4 targeted
validation phase** of the NSTC grant.

---

# Outputs

| File | Description |
|------|-------------|
| `output/tables/Table3_enriched_pathways.csv` | Pathway enrichment (manuscript Table 3) |
| `output/tables/Table_sidedness_differential.csv` | Right-vs-left tumour per-feature results |
| `output/tables/Table_biomarker_panel.csv` | Top 20 biomarker candidates |
| `output/figures/06_Pathway_*` | Pathway enrichment figures (Figure 6) |
| `output/figures/FigS*_sidedness_*` | Sidedness figures (supplementary) |
| `output/figures/Figure_08_ROC_biomarker_panel.*` | ROC figure (Figure 8) |

---

# Session info

```{r session-info}
sessionInfo()
```
